\bf Mainstreaming Race Science

### 2023-07-10 paper/paper.Rmd
enable cross-references

---
title: "Mainstreaming Race Science"
authors:
  - name: "Daniel J. Hicks"
    affiliation: 1
    corresponding_author: true
  - name: "Emilio Lobato"
    affiliation: 1
address:
  - code: 1
    address: "University of California, Merced"
corresponding_author: "Corresponding author: Daniel J. Hicks, dhicks4@ucmerced.edu."
abstract: |
  *[structured]*
acknowledgements: |
  *[strutured]* Thanks to Emily Merchant and John Jackson for some initial discussions that helped clarify the scope of the project and identify some key resources related to the Pioneer Fund and *Mankind Quarterly*.  Thanks to Anthony Sainez for help retrieving the *Mankind Quarterly* articles.  Thanks to Derek Devnich and James Dooley for their work attempting to secure the APA-published articles.  EL's work on this project was supported by UC Merced.  DH's work on this project was not supported by any specific funding. 
bibliography: "references.bib"
output: 
  # https://github.com/rstudio/rticles/pull/486
  # skeleton: <https://github.com/christopherkenny/rticles/tree/master/inst/rmarkdown/templates/science>
  rticles::science_article:
      keep_tex: true
      move_figures: false
      draft: false
      number_sections: false
---

@@

*[for xref: https://bookdown.org/yihui/bookdown/a-single-document.html]*

We identified 16 researchers who had received funding from Pioneer; 14 of these researchers had profiles in the Web of Science (WoS) author search *[table + WoS pub coucnt]*, allowing us identify 13 WoS-indexed journals that had published 6 or more of these authors.

@@

Table: Pioneer-funded researchers.  Either identified in *[cite]* or named on Pioneer's website, along with identified discipline and WoS author search result counts.

![Journals publishing 6 or more Pioneer-funded researchers, Web of Science author search results *[date coverage]*](img/wos_results.png){width="4.76in" height="2.6in"}

Figure *[fig]* shows that, while MQ is among the "Pioneer-publishing" journals, a number of mainstream journals are more prominent: *Personality and Individual Differences* (PID), *Intelligence* (I), *Behavior Genetics* (BG), and *Psychological Reports* (PR).  In addition, only psychologist Richard Lynn appears to have published heavily in MQ.  Lynn has been an assistant editor of MQ since 1979 *[check this: https://rationalwiki.org/wiki/Mankind_Quarterly]* and president of PF since the death of psychologist J. Phillippe Rushton in 2012 *[cite]*.  By contrast, a number of Pioneer-funded researchers have published in PID, I, and to a lesser degree BG: Bouchard, Eysenck, Jensen, Rushton, Vernon, and also Lynn.

@@

The topic model analysis and the case of Whitney's address suggest that, by the 1990s, behavior genetics had at least somewhat distanced itself from race science.  *[https://link.springer.com/search?use-oscar-shared-search=true&sortBy=newestFirst&dateTo=2010&sortOrder=newestFirst&facet-journal-id=10519&content-type=Article&query=race+intelligence&date=custom&dateFrom=&date-facet-mode=between&facet-start-year=1972&previous-start-year=1972&facet-end-year=2010&previous-end-year=2023]*

A disciplinary history of behavior genetics published in 2010 \cite{McGueEndBehavioralGenetics2010} acknowledges the historical association with eugenics, but does not mention Jensen and other race science controversies since World War II. Instead postwar behavior genetics is portrayed as an endeavor to counterbalance the hegemony of the "Blank Slate," the supposedly widespread view that genetic differences play no role in explaining individual differences \cite{BatesonCorpseWearisomeDebate2002}.

These findings for *Behavior Genetics* complicate at least the popular association of behavior genetics with race science, if not Panofsky's more nuanced analysis.

@@

*[limitation: colorblind racism (Bonilla-Silva) creates coding ambiguities]*

>>>>

---
title: "Mainstreaming Race Science"
authors:
  - name: "Daniel J. Hicks"
    affiliation: 1
    corresponding_author: true
  - name: "Emilio Lobato"
    affiliation: 1
address:
  - code: 1
    address: "University of California, Merced"
corresponding_author: "Corresponding author: Daniel J. Hicks, dhicks4@ucmerced.edu."
abstract: |
  *[structured]*
acknowledgements: |
  *[strutured]* Thanks to Emily Merchant and John Jackson for some initial discussions that helped clarify the scope of the project and identify some key resources related to the Pioneer Fund and *Mankind Quarterly*.  Thanks to Anthony Sainez for help retrieving the *Mankind Quarterly* articles.  Thanks to Derek Devnich and James Dooley for their work attempting to secure the APA-published articles.  EL's work on this project was supported by UC Merced.  DH's work on this project was not supported by any specific funding. 
bibliography: "references.bib"
output: 
  # https://github.com/rstudio/rticles/pull/486
  # skeleton: <https://github.com/christopherkenny/rticles/tree/master/inst/rmarkdown/templates/science>
  # rticles::science_article:
  #   keep_tex: true
  #    move_figures: false
  #    draft: false
  #    number_sections: false
  bookdown::pdf_book:
    base_format: rticles::science_article
    keep_tex: true
    move_figures: false
    draft: false
    number_sections: false
---

@@

We identified 16 researchers who had received funding from Pioneer; 14 of these researchers had profiles in the Web of Science (WoS) author search *[table + WoS pub coucnt]*, allowing us identify 13 WoS-indexed journals that had published 6 or more of these authors. See Table \@ref(tab:researchers).

@@

Table: (\#tab:researchers) Pioneer-funded researchers.  Either identified in *[cite]* or named on Pioneer's website, along with identified discipline and WoS author search result counts.

![Journals publishing 6 or more Pioneer-funded researchers, Web of Science author search results *[date coverage]* (\#fig:wos)](img/wos_results.png){width="4.76in" height="2.6in"}

Figure \@ref(fig:wos) shows that, while MQ is among the "Pioneer-publishing" journals, a number of mainstream journals are more prominent: *Personality and Individual Differences* (PID), *Intelligence* (I), *Behavior Genetics* (BG), and *Psychological Reports* (PR).  In addition, only psychologist Richard Lynn appears to have published heavily in MQ.  Lynn has been an assistant editor of MQ since 1979 *[check this: https://rationalwiki.org/wiki/Mankind_Quarterly]* and president of PF since the death of psychologist J. Phillippe Rushton in 2012 *[cite]*.  By contrast, a number of Pioneer-funded researchers have published in PID, I, and to a lesser degree BG: Bouchard, Eysenck, Jensen, Rushton, Vernon, and also Lynn.

@@

The topic model analysis and the case of Whitney's address suggest that, by the 1990s, behavior genetics had at least somewhat distanced itself from race science, though perhaps without repudiating it.  *[https://link.springer.com/search?use-oscar-shared-search=true&sortBy=newestFirst&dateTo=2010&sortOrder=newestFirst&facet-journal-id=10519&content-type=Article&query=race+intelligence&date=custom&dateFrom=&date-facet-mode=between&facet-start-year=1972&previous-start-year=1972&facet-end-year=2010&previous-end-year=2023]*

To check this interpretation, we ran an independent search for documents with the keywords "race" and "intelligence" published in BG from 1972 to 2020, using Springer's journal search website.  This search returned 125 documents.  33 were presentation abstracts from meetings of the Behavior Genetics Association; we did not examine these further.  14 of the remaining documents were published since 2010 (after the scope of the primary study).  Among these, one obituary and two papers celebrated the work of John Loehlin, a prominent race-and-intelligence researcher who had been Director of the American Eugenics Society from 1968-1972 *[http://link.springer.com/article/10.1007/s10519-020-10024-w; http://link.springer.com/article/10.1007/s10519-014-9687-1; http://link.springer.com/article/10.1007/s10519-014-9673-7]*.  Another was a retrospective of the work of Lindon Eaves, a geneticist who made at least one notable contribution to debates on the Scarr-Rowe hypothesis (interactions between heritability, race, and class).  None of these 14 documents reported any studies of racial differences.

28 documents in this sample were published between 1990 and 2010.  One was an obituary of Jerry Hirsch, a critic of hereditarianism in general and Jensen in particular *[http://link.springer.com/article/10.1007/s10519-008-9231-2]*, and another was an obituary of David Rowe *[http://link.springer.com/article/10.1023/A%3A1026122928868]*.  Only 2 of these 28 documents reported race differences of any kind: one examining at interactions among race, sex, and heritability for adolescent BMI (Body Mass Index, used as a measure of overweight/obesity) *[http://link.springer.com/article/10.1023/A%3A1021619329904]*; and the other at interactions among race, family history of alcoholism, and visuospatial performance *[http://link.springer.com/article/10.1007/BF02197241]*.  Finally, Philip Vernon — one of the Pioneer-funded scientists we identified earlier — published a critical review of a book that attempted to address hereditarianism, the Flynn effect, race-and-intelligence research, and some related issues.

This supplemental review provides additional support for the interpretation that behavior genetics had distanced itself from race science by the 1990s.  Notably, this review brought our attention to a disciplinary history of behavior genetics published in 2010 \cite{McGueEndBehavioralGenetics2010}, which acknowledges the historical association with eugenics but does not mention Jensen and other race science controversies since World War II. Instead postwar behavior genetics is portrayed as an endeavor to counterbalance the hegemony of the "Blank Slate," the supposedly widespread view that genetic differences play no role in explaining individual differences \cite{BatesonCorpseWearisomeDebate2002}. And none of the three documents celebrating Loehlin noted his postwar involvement in the eugenics movement.  None of the documents examined in this supplemental review explicitly repudiate race science.  Behavior geneticists seem to typically adopt a posture of objectivity and technical rigor, downplaying the extent of the field's involvement in race science in the past.

@@

*[limitation: colorblind racism (Bonilla-Silva; WinstonWhyMainstreamResearch2020) creates coding ambiguities]*

### 2023-07-10 paper/paper.tex
enable cross-references

\emph{{[}incorporate def'ns{]}} scientific racism \textasciitilde{}
purporting to justify racial inequality and colonialism by appealing to
the epistemic authority of science

\begin{description}
\tightlist
\item[race science]
(pseudo-)scientific research that can be utilized for scientific racism
\item[race science discourse]
treating race science as a legitimate area of scientific research; this
includes methodological critiques of race science and empirical tests
that falsify race science hypotheses
\end{description}

@@

\emph{{[}for xref:
\url{https://bookdown.org/yihui/bookdown/a-single-document.html}{]}}

We identified 16 researchers who had received funding from Pioneer; 14
of these researchers had profiles in the Web of Science (WoS) author
search \emph{{[}table + WoS pub coucnt{]}}, allowing us identify 13
WoS-indexed journals that had published 6 or more of these authors.

\begin{longtable}[]{@{}lrl@{}}
\caption{Pioneer-funded researchers. Either identified in
\emph{{[}cite{]}} or named on Pioneer's website, along with identified
discipline and WoS author search result counts.}\tabularnewline
\toprule
\endhead
Thomas J. Bouchard, Jr. & psychology & \\
Brunetto Chiarelli & anthropology & \\
Hans Eysenck & psychology & \\
Robert Gordon & sociology & \\
Linda Gottfredson & psychology & \\
Garrett Hardin & ecology & \\
Joseph M. Horn & psychology & \\
Lloyd Humphreys & psychology & \\
Arthur Jensen & psychology & \\
Michael Levin & philosophy & \\
Richard Lynn & psychology & \\
R. Travis Osborne & psychology & \\
J. Phillippe Rushton & psychology & \\
Audrey M. Shuey & psychology & \\
Philip A. Vernon & psychology & \\
Daniel Vining, Jr. & demography & \\
\bottomrule
\end{longtable}

\begin{figure}
\centering
\includegraphics[width=4.76in,height=2.6in]{img/wos_results.png}
\caption{Journals publishing 6 or more Pioneer-funded researchers, Web
of Science author search results \emph{{[}date coverage{]}}}
\end{figure}

Figure \emph{{[}fig{]}} shows that, while MQ is among the
``Pioneer-publishing'' journals, a number of mainstream journals are
more prominent: \emph{Personality and Individual Differences} (PID),
\emph{Intelligence} (I), \emph{Behavior Genetics} (BG), and
\emph{Psychological Reports} (PR). In addition, only psychologist
Richard Lynn appears to have published heavily in MQ. Lynn has been an
assistant editor of MQ since 1979 \emph{{[}check this:
\url{https://rationalwiki.org/wiki/Mankind_Quarterly}{]}} and president
of PF since the death of psychologist J. Phillippe Rushton in 2012
\emph{{[}cite{]}}. By contrast, a number of Pioneer-funded researchers
have published in PID, I, and to a lesser degree BG: Bouchard, Eysenck,
Jensen, Rushton, Vernon, and also Lynn.

@@

The topic model analysis and the case of Whitney's address suggest that,
by the 1990s, behavior genetics had at least somewhat distanced itself
from race science.
\emph{{[}\url{https://link.springer.com/search?use-oscar-shared-search=true\&sortBy=newestFirst\&dateTo=2010\&sortOrder=newestFirst\&facet-journal-id=10519\&content-type=Article\&query=race+intelligence\&date=custom\&dateFrom=\&date-facet-mode=between\&facet-start-year=1972\&previous-start-year=1972\&facet-end-year=2010\&previous-end-year=2023}{]}}

A disciplinary history of behavior genetics published in 2010
\cite{McGueEndBehavioralGenetics2010} acknowledges the historical
association with eugenics, but does not mention Jensen and other race
science controversies since World War II. Instead postwar behavior
genetics is portrayed as an endeavor to counterbalance the hegemony of
the ``Blank Slate,'' the supposedly widespread view that genetic
differences play no role in explaining individual differences
\cite{BatesonCorpseWearisomeDebate2002}.

These findings for \emph{Behavior Genetics} complicate at least the
popular association of behavior genetics with race science, if not
Panofsky's more nuanced analysis.

@@

\emph{{[}this probably goes too far even for a discussion section{]}}

@@

>>>>

\emph{{[}incorporate def'ns{]}}

\begin{description}
\tightlist
\item[scientific racism]
purporting to justify racial inequality and colonialism by appealing to the epistemic authority of science
\item[race science]
(pseudo-)scientific research that can be utilized for scientific racism
\item[race science discourse]
treating race science as a legitimate area of scientific research; this includes methodological critiques of race science and empirical tests that falsify race science hypotheses
\end{description}

\begin{itemize}
\tightlist
\item
  pre-20th century history of race concepts and race science
\item
  expansion and contraction of racialized terms over time
\end{itemize}

@@

We identified 16 researchers who had received funding from Pioneer; 14 of these researchers had profiles in the Web of Science (WoS) author search \emph{{[}table + WoS pub coucnt{]}}, allowing us identify 13 WoS-indexed journals that had published 6 or more of these authors. See Table \ref{tab:researchers}.

\begin{longtable}[]{@{}lrl@{}}
\caption{\label{tab:researchers} Pioneer-funded researchers. Either identified in \emph{{[}cite{]}} or named on Pioneer's website, along with identified discipline and WoS author search result counts.}\tabularnewline
\toprule
\endhead
Thomas J. Bouchard, Jr. & psychology & \\
Brunetto Chiarelli & anthropology & \\
Hans Eysenck & psychology & \\
Robert Gordon & sociology & \\
Linda Gottfredson & psychology & \\
Garrett Hardin & ecology & \\
Joseph M. Horn & psychology & \\
Lloyd Humphreys & psychology & \\
Arthur Jensen & psychology & \\
Michael Levin & philosophy & \\
Richard Lynn & psychology & \\
R. Travis Osborne & psychology & \\
J. Phillippe Rushton & psychology & \\
Audrey M. Shuey & psychology & \\
Philip A. Vernon & psychology & \\
Daniel Vining, Jr. & demography & \\
\bottomrule
\end{longtable}

\begin{figure}
\centering
\includegraphics[width=4.76in,height=2.6in]{img/wos_results.png}
\caption{Journals publishing 6 or more Pioneer-funded researchers, Web of Science author search results \emph{{[}date coverage{]}} \label{fig:wos}}
\end{figure}

Figure \ref{fig:wos} shows that, while MQ is among the ``Pioneer-publishing'' journals, a number of mainstream journals are more prominent: \emph{Personality and Individual Differences} (PID), \emph{Intelligence} (I), \emph{Behavior Genetics} (BG), and \emph{Psychological Reports} (PR). In addition, only psychologist Richard Lynn appears to have published heavily in MQ. Lynn has been an assistant editor of MQ since 1979 \emph{{[}check this: \url{https://rationalwiki.org/wiki/Mankind_Quarterly}{]}} and president of PF since the death of psychologist J. Phillippe Rushton in 2012 \emph{{[}cite{]}}. By contrast, a number of Pioneer-funded researchers have published in PID, I, and to a lesser degree BG: Bouchard, Eysenck, Jensen, Rushton, Vernon, and also Lynn.

@@

The topic model analysis and the case of Whitney's address suggest that, by the 1990s, behavior genetics had at least somewhat distanced itself from race science, though perhaps without repudiating it. \emph{{[}\url{https://link.springer.com/search?use-oscar-shared-search=true\&sortBy=newestFirst\&dateTo=2010\&sortOrder=newestFirst\&facet-journal-id=10519\&content-type=Article\&query=race+intelligence\&date=custom\&dateFrom=\&date-facet-mode=between\&facet-start-year=1972\&previous-start-year=1972\&facet-end-year=2010\&previous-end-year=2023}{]}}

To check this interpretation, we ran an independent search for documents with the keywords ``race'' and ``intelligence'' published in BG from 1972 to 2020, using Springer's journal search website. This search returned 125 documents. 33 were presentation abstracts from meetings of the Behavior Genetics Association; we did not examine these further. 14 of the remaining documents were published since 2010 (after the scope of the primary study). Among these, one obituary and two papers celebrated the work of John Loehlin, a prominent race-and-intelligence researcher who had been Director of the American Eugenics Society from 1968-1972 \emph{{[}\url{http://link.springer.com/article/10.1007/s10519-020-10024-w}; \url{http://link.springer.com/article/10.1007/s10519-014-9687-1}; \url{http://link.springer.com/article/10.1007/s10519-014-9673-7}{]}}. Another was a retrospective of the work of Lindon Eaves, a geneticist who made at least one notable contribution to debates on the Scarr-Rowe hypothesis (interactions between heritability, race, and class). None of these 14 documents reported any studies of racial differences.

28 documents in this sample were published between 1990 and 2010. One was an obituary of Jerry Hirsch, a critic of hereditarianism in general and Jensen in particular \emph{{[}\url{http://link.springer.com/article/10.1007/s10519-008-9231-2}{]}}, and another was an obituary of David Rowe \emph{{[}\url{http://link.springer.com/article/10.1023/A\%3A1026122928868}{]}}. Only 2 of these 28 documents reported race differences of any kind: one examining at interactions among race, sex, and heritability for adolescent BMI (Body Mass Index, used as a measure of overweight/obesity) \emph{{[}\url{http://link.springer.com/article/10.1023/A\%3A1021619329904}{]}}; and the other at interactions among race, family history of alcoholism, and visuospatial performance \emph{{[}\url{http://link.springer.com/article/10.1007/BF02197241}{]}}. Finally, Philip Vernon --- one of the Pioneer-funded scientists we identified earlier --- published a critical review of a book that attempted to address hereditarianism, the Flynn effect, race-and-intelligence research, and some related issues.

This supplemental review provides additional support for the interpretation that behavior genetics had distanced itself from race science by the 1990s. Notably, this review brought our attention to a disciplinary history of behavior genetics published in 2010 \cite{McGueEndBehavioralGenetics2010}, which acknowledges the historical association with eugenics but does not mention Jensen and other race science controversies since World War II. Instead postwar behavior genetics is portrayed as an endeavor to counterbalance the hegemony of the ``Blank Slate,'' the supposedly widespread view that genetic differences play no role in explaining individual differences \cite{BatesonCorpseWearisomeDebate2002}. And none of the three documents celebrating Loehlin noted his postwar involvement in the eugenics movement. None of the documents examined in this supplemental review explicitly repudiate race science. Behavior geneticists seem to typically adopt a posture of objectivity and technical rigor, downplaying the extent of the field's involvement in race science in the past.

@@

\emph{{[}limitation: colorblind racism (Bonilla-Silva; WinstonWhyMainstreamResearch2020) creates coding ambiguities{]}}

\emph{{[}limitation: false negatives (race science papers that don't appear in V24) are possible if, eg, they use many terms related to twin studies of heritability and thereby appear in V03{]}}

\emph{{[}replace w/: WinstonWhyMainstreamResearch2020: race science as a community project of psychology(?). radical academic freedom is a tool used to defend race science PigliucciWhatAreWe2013, RosemanTroublesomeReflectionRacism2014, LarsenMoreProvocativeLess2020. an important part of how these mainstream journals function as race science echo chambers.{]}}

@@

\emph{{[}what can science do? Posselt, /Equity in Science/{]}}
